Trauma treatment delay

### 2023-09-20 lab-book.Rmd
Lekt med dataset
Trauma system intro
Introduction
intro and code

inga barn med (15 under)
delay är OFI. vilka patienter?
traumaregister kopplat till swetrau, inte samma dock. swetrau, NISS 15, NKS ESS 9 b.l.a.
kvalitetsregister är det konferenser sammanställer. auditfilters faller där, kallar till konferens om OFI är positivt
create OFI, skapar binärt ja eller nej
kolum i dataset som heter DT_treatment typ. delta time. ph prehospital. ed emergency department
kolum är tiden till första hjälpande insats. kolum är i minuter, kom till hjälp.
olika faktorer som höjer tid.
DT treatment finns ej på 43% OFI, 74% av ej OFI.

>>>>

inga barn med (15 under)
delay är OFI. vilka patienter?
traumaregister kopplat till swetrau, inte samma dock. swetrau, NISS 15, NKS ESS 9 b.l.a.
kvalitetsregister är det konferenser sammanställer. auditfilters faller där, kallar till konferens om OFI är positivt
create OFI, skapar binärt ja eller nej.
typ. delta time. ph prehospital. ed emergency department
kolum är tiden till första hjälpande insats. kolum är i minuter, kom till hjälp.
olika faktorer som höjer tid.
DT treatment finns ej på 43% OFI, 74% av ej OFI.
dt_ed_emerg_prod är min

mycket data finns på patientfaktorer o mortalitet, men ej fördröjning

konferenser ovalninga globalt, resurskrävande. bara KS i sverige. nästan alla sjukhus går igenom dödsfall.
Finns på KI 2010-2012. hjärtstopp ung man, vet ej hur man skulle göra. intresekonflikt mellan medicin o kirurg.
gunilla wilke? anestesisköterska styrde upp mha två läkare kvalitetsförbättringsarbete. 2013 började? 
gick först igenom dödsfall, sen utanför mortalitet. försökte inkorporera audit filters, mellan 2013-2017 audit filter för brett. valde patienter på måfå.
gjorde om filter till de vi har nu, fick in en till sjuksyrra. liselott applicerar alla filter, typ 9 olika. om ja, går igenom pat med gunilla.
på konferens alla som sysslar med trauma

plan B, vi har all data så det ska gå

tog bort
csl: bmcemerg.csl

### 2023-09-20 manuscriptlongversion.Rmd
intro and code

Within the realm of modern medicine, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients. The definition of medical trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020]. As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological. The one we will be focusing on in this paper is physical trauma, which is defined as a serious injury to the body. It is often divided into two subgroups, blunt force trauma and penetrative trauma. Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts. Examples of this could be car crashes, falls or direct blows to the body. Penetrative trauma is when an object pierces the skin or body and creates an open wound. Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020]. Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale on arrival and as such require more resources and be hospitalized for a longer period because of their more sever injuries[@FITCH20191201].

@@

@@

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012]. During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%. This is attributed to the improvements in care for acute brain injuries and bleeding control. The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008]. Functioning trauma systems saves lives in the hospital, but their work outside of the hospital is just as important. Data collection from injured patients, such as mechanism of injury or mortality, are essential for creating databases that can be used for research. In turn, that research can be used for planning injury prevention programs that target the most common injuries in the right ways, e.g. teen drivers, children, specific occupations etc. These injury prevention programs can be planned on trauma center, organizational or government level[@ACS].

Trauma teams are multidisciplinary and operate in these centers. They play a pivotal role in the initial care of the trauma patient, as they are primary care givers in the critical stage of trauma. In Sweden, they consist of individuals from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader. Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care. Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel. Once in the trauma room, the team works through ABCDE, a systematic and structured method to assess and manage critically ill patients. It is designed to handle the most urgent problems first, such as airways and breathing, and then go through the rest of the body in a structured way so to not miss anything. [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].

@@

Once treated, the patients information and timeline of what happened enters a trauma care database, depending on what hospital. At Nya Karolinska Universitetssjukhuset, all patients who have been treated and had ISS\>9 or had the trauma alarm activated are included and screened for OFIs. A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs and implement corrective actions. This refers to a meeting where doctors from different specialities sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database. Examples of opportunities for improvement may include lack of resources and management errors. One common OFI that this paper will look further into is delayed treatment. Among the preventable or possibly preventable deaths in trauma patients, a subgroup of patients determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007]. Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood. There may be several factors that correlate with receiving delayed treatment, but they have yet to be identified. Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.

@@

This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. The trauma registry includes about 12000 patients treated between 2012 and 2022. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022. The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database. This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression. A 5% significance level and 95% confidence levels will be used.

>>>>

Within modern hospital care, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients. The definition of trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020]. As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological. The one we will be focusing on in this paper is physical trauma, the clinical entity composed of physical injury and the body's associated response, hereby referred to as trauma.

@@

It is often divided into two subgroups, blunt force trauma and penetrative trauma. Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts. Examples of this could be car crashes, falls or direct blows to the body. Penetrative trauma is when an object pierces the skin or body and creates an open wound. Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020]. Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale on arrival and as such require more resources and be hospitalized for a longer period because of their more sever injuries.

@@

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@Celso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012]. During the last three decades, the introduction of trauma systems has contributed to lowering the incidence of preventable death . This is attributed to the improvements in care for acute brain injuries and bleeding control. The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008]. Functioning trauma systems saves lives in the hospital, but their work outside of the hospital is just as important. Data collection from injured patients, such as mechanism of injury or mortality, are essential for creating databases that can be used for research. In turn, that research can be used for planning injury prevention programs that target the most common injuries in the right ways, e.g. teen drivers, children, specific occupations etc. These injury prevention programs can be planned on trauma center, organizational or government level[@ACS].

Trauma teams are multidisciplinary and operate in these centers. They play a pivotal role in the initial care of the trauma patient, as they are primary care givers in the critical stage of trauma. In Sweden, they consist of individuals from the specialities of surgery, intensive care, orthopedics, nursing and support staff and are lead by a team leader. Their aim is to rapidly assess and stabilise the patient, prioritise their injuries and arrange for site of definitive care. Emergency nurses are often the personnel activating the trauma code using anatomical and physiological criteria, mechanism of injury and information obtained from emergency service personnel. Once in the trauma room, the team works systematically to assess and manage critically ill patients. They handle the most urgent problems first, such as airways and breathing, and then tend to less life threatening injuries before readying the patient for transport to the next site of care [@Hedberg_Falkén_Wihlke_Jansson_Hultgren_Eriksson_Beckman_Haupt_Kjörholt_Sandersjöö_et_al._2020].

@@

Many hospitals, mainly in HIC, have a registry database where they log the patients information and timeline of what happened. A cornerstone of trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify OFIs and implement corrective actions. This refers to a meeting where different specialities who work with trauma care sit down and go through cases where possible OFIs were identified by nurses who go through the trauma care quality database. Examples of opportunities for improvement may include lack of resources and management errors. One common OFI that this paper will look further into is delayed treatment. Among the preventable or possibly preventable deaths in trauma patients, a subgroup of patients determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007]. Although it is such a common OFI, the patient level factors associated with delayed treatment are poorly understood. There may be several factors that correlate with receiving delayed treatment, but they have yet to be identified. Mapping these factors may help in identifying patients that might be at risk for receiving delayed treatment before it happens.

@@

This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. The trauma registry includes about 12000 patients treated between 2012 and 2022. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2013 and 2022. The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database. This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression. A 5% significance level and 95% confidence levels will be used.
